\begin{abstract}

Current microprocessors are limited by power density and new designs must emphasize
energy efficiency to become successful. Building energy efficient hardware
requires better understanding of how the ISA and its implementation relates to
energy efficiency. This paper investigates the instruction level energy
efficiency of an ARM Cortex-A9, and presents numbers on current drain, pipeline
utilization and other performance data for different instructions. This allows
reasoning about how instructions are executed and how energy efficient their
implementation are in this otherwise publicly closed architecture.

The testbench used in the experiments measures the core current drain by
utilizing a shunt resistor connected directly between the CPU core voltage
supply pins and an external power supply. Measuring isolated core current drain
effectively means skipping other on- and off-chip peripherals as much as
possible, and get a more decent view of what is going on within the processor
core.  We exploit instruction buffers within the processor and avoid fetching
data from memory. This renders the memory hierarchy unused, and isolates the
core as much as possible.

The results shows that the ARM ISA is well balanced regarding commonly used
instructions, with instructions such as \texttt{add} and \texttt{sub} seemingly
most energy efficient. Unfortunately the ISA suffers from an inefficient handling
of conditional execution and status flags. Both terms cause what is assumed to
be forced synchronization, which again forces inefficient utilization of the
otherwise computationally strong core.

\end{abstract}
